`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-02-11 paper/paper.qmd
typos and polish

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [tierney2018novel], setting rehabilitation districts for stroke patients [padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].

@@

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017]. To provide easy access to the existing friction surfaces generated by @weiss2020global. `traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent. Users can also supply other friction surfaces. Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.

@@

Other R packages provide superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

With `traveltime`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

@@

In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.

## Prepare the data and download friction surface

@@

We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata]. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`). Our boundary `singapore_shapefile` is a `SpatVector` class object.

@@

Now that we have the two items of data that we require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `singapore_shapefile`, that are within the extent of the friction surface raster:

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}). `traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell. In this case walking friction is fairly uniform across the islands barring a few small inaccesible locations. More heterogeneity would be expected in less developed areas with fewer transport corridors.

@@

With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore()` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:

@@

We present the resulting calculated travel times in Figure \ref{fig-result}. Note that the results in `trave_time_singapore` include infinite (`Inf`) values. In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. The raster cells for these islands appear absent in Figure \ref{fig-result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.

@@

Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2]. Although the user can supply their own friction surface, we expect most applications will use these surfaces. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell.

>>>>

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].

@@

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` class object; and returns the result as a `SpatRaster` class object. The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017]. To provide easy access to the existing friction surfaces generated by @weiss2020global, `traveltime` uses the package `malariaAtlas` to allow users to access friction surfaces for the area of interest; though users can also supply any other friction surfaces to `traveltime`. Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying methodological work [@weiss2018global; @weiss2020global] as well.

@@

Other R packages provide superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location. The isochron polygons are most comparable to what `traveltime::calculate_travel_time()` calculates, though each isochron polygon is a single polygon calculated is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than a raster of gradation across a landscape for an arbitrary number of points as in `traveltime`. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

With `traveltime`, we provide free and open source software to estimate travel time from any number of user-supplied locations, across a complete area of interest, and with convenient access to motorised or walking friction surfaces with global coverage.

@@

In this example we will calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.

## Prepare the data and friction surface

@@

We can download a national-level polygon boundary of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata]. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`). Our boundary `singapore_shapefile` is a `SpatVector` class object.

@@

Now that we have the two items of data that we require, the next step is to prepare a friction surface for our area of interest. We will use the friction surface from @weiss2020 that can be downloaded by `traveltime` with the function `get_friction_surface()`. We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of the land boundary in `singapore_shapefile`:

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points (Figure \ref{fig-data}). `traveltime` takes resistance values of friction [@gdistance2017], so higher values of friction indicate more time travelling across a given cell. In this case walking friction is fairly uniform across the islands barring a few small inaccesible locations. More heterogeneity would be expected in less developed locations with fewer transport corridors.

@@

With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore` and the points of interest in `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:

@@

We present the resulting calculated travel times in Figure \ref{fig-result} where, as expected, the travel times are lowest near station exits (per Figure \ref{fig-data}) and progressively higher further away. Note that the results in `trave_time_singapore` include infinite values (`Inf`  above). In Figure \ref{fig-data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. The raster cells for these islands appear absent in Figure \ref{fig-result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig-result}.

@@

Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2]. Although the user can supply their own friction surface, we expect most applications will use these existing surfaces given the extensive work needed in creating new ones [@weiss2018global; @weiss2020global]. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km^2^ cell size is a relatively large area to walk across, and thus actual waking times may vary significantly within each cell. We underline however that a user can provide their own higher resolution friction surface at present if desired.
